---
title: "Merging BrainChart and ENIGMA outputs"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Install Packages
```{r Install new packages}
# corr, ggcorrplot, and FactoMineR are packages used for performing principal component analysis (PCA) 
# install.packages("corrr")
# install.packages("ggcorrplot")
# install.packages("FactoMineR") # unable to install - ERROR: dependency ‘emmeans’ is not available for package ‘FactoMineR’
# install.packages("emmeans")
# install.packages("evaluate")

# install.packages("bookdown") -> may need to re-install before knitting

```


## Load Libraries
```{r Load libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)
# library(corrr)
# library(ggcorrplot)
# library("FactoMineR")

```

## Load data
```{r Read csv files as data frames}

# Load csv with brain age gap calculated using ENIGMA age prediction output
enigma_new <- read_csv("/external/rprshnas01/tigrlab/scratch/csun/brainage_proj/data/processed/enigma_new.csv")
# Note the version of total disorder groupings used in enigma_new (0-1, 2-3, 4+) 
enigma_new
# Load csv with centile scores from BrainChart
centiles_final <- read_csv("/KIMEL/tigrlab/scratch/csun/BrainChart_data/data/processed/centiles_final.csv")
# Note the version of total disorder groupings used in centiles_final (0-1, 2-3, 4+) 

# Load csv with enigma csv with ASD diagnoses appended
enigma_ASD <- read_csv("/external/rprshnas01/tigrlab/scratch/csun/brainage_proj/data/processed/enigma_ASD.csv") 

# load group_T1w.tsv 
group_T1w <- read_delim("/external/rprshnas01/tigrlab/archive/data/TAY/pipelines/versioned_outputs/baseline_only/mriqc/group_T1w.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
group_T1w

# load mriqc QA data for participants included in the protocol paper
# this version of QA data is for reference only, group_T1w contains a more recent version of qi_1 values (i.e., more participants are included)
# TAY_mriqc <- read_csv("/KIMEL/tigrlab/scratch/csun/brainage_proj/data/raw/mriqc_tay_group_T1w_20220414.csv")
# TAY_mriqc

# load T1w tsv file from jwong for sub-CMH00000263_ses-01_T1w
sub_263_T1w <- read_delim("/KIMEL/tigrlab/scratch/csun/brainage_proj/data/raw/group_T1w_og.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
sub_263_T1w

# load T1w tsv file from jwong for sub-CMH00000433 T1w
sub_433_T1w <- read_delim("/KIMEL/tigrlab/scratch/jwong/mriqc_sub433/output/group_T1w.tsv", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
sub_433_T1w

# append participant 433's T1w IQMs to end of group_T1w df
group_T1w <- rbind(group_T1w, sub_433_T1w)

# load group_bold.tsv - NOT NECESSARY for T1w IQMs
group_bold <- read_delim("/external/rprshnas01/tigrlab/archive/data/TAY/pipelines/versioned_outputs/baseline_only/mriqc/group_bold.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
group_bold
```
Comments on the qi_1 values: 
- because subject 433 has a qi_1 value of 0, in the context of graphs and implications for subsequent analyses, this participant will not have much of an influence on results (we are mostly interested in participants with non-zero qi_1 values)

## Incorporate QA metrics
Resource with tips and tricks for using the dplyr filter function: https://sebastiansauer.github.io/dplyr_filter/ 
```{r Add qi_1 data}
# filter out participants with zero values for qi_1
group_T1w %>% 
  select(bids_name, qi_1) %>% # include only these 2 columns 
  filter(qi_1 != 0) # filter for rows that have non-zero qi_1 values  
# There are 49 entries of non-zero qi_1 values but some of these correspond to multiple runs for the same participant

# filter for participants with multiple runs
multi_run <- group_T1w %>% 
  select(bids_name, qi_1) %>% 
  filter(str_detect(bids_name, "run")) # filter for rows with multiple runs
# There are 36 entries corresponding to participants with multiple runs, including some that have qi_1 values of zero for both runs.
multi_run

# filter for participants with non-zero qi_1 AND multiple runs
group_T1w %>% 
  select(bids_name, qi_1) %>% # include only these 2 columns 
  filter(qi_1 != 0) %>% # filter for rows that have non-zero qi_1 values  
  filter(str_detect(bids_name, "run")) # filter for rows with multiple runs
# There are 15 entries of non-zero qi_1 values corresponding to multiple runs

# save data frame w/ participants who have multiple runs as csv for reference purposes
write.csv(multi_run, "/KIMEL/tigrlab/scratch/csun/brainage_proj/data/raw/multi_run.csv", row.names = FALSE)
```

Assumptions for removing and merging qi_1 values for participants with multiple runs -- double checked with Erin and Jimmy
- If one of the runs is a blacklisted scan, remove it and keep the remaining scan
- If one run is flagged while the other is approved, remove the flagged scan and keep the approved
- If both runs are flagged, keep the run with the smaller qi_1 value (?)
- If both runs have qi_1 value of 0, merge and keep the qi_1 value at 0
- 
```{r Filtering qi_1 data}
#### remove qi_1 values for blacklisted scans ####
blacklist <- c("sub-CMH00000218_ses-01_run-01_T1w", "sub-CMH00000224_ses-01_run-02_T1w", "sub-CMH00000283_ses-01_run-01_T1w", "sub-CMH00000421_ses-01_run-01_T1w", "sub-CMH00000466_ses-01_run-01_T1w")
class(blacklist)
blacklist_T1 <- group_T1w[group_T1w$bids_name %in% c("sub-CMH00000218_ses-01_run-01_T1w", "sub-CMH00000224_ses-01_run-02_T1w", "sub-CMH00000283_ses-01_run-01_T1w", "sub-CMH00000421_ses-01_run-01_T1w", "sub-CMH00000466_ses-01_run-01_T1w"), ]
blacklist_T1

#### remove qi_1 values for flagged runs ####
# includes flagged runs that are paired with an approved run and flagged runs with larger qi_1 
flagged_T1 <- group_T1w[group_T1w$bids_name %in% c("sub-CMH00000062_ses-01_run-02_T1w", "sub-CMH00000150_ses-01_run-01_T1w","sub-CMH00000192_ses-01_run-01_T1w", "sub-CMH00000263_ses-01_run-01_T1w", "sub-CMH00000360_ses-01_run-01_T1w", "sub-CMH00000361_ses-01_run-01_T1w","sub-CMH00000369_ses-01_run-01_T1w"), ]
flagged_T1

#### runs that both have qi_1 values of 0 ####
# delete one of the rows, choose the one that is flagged if the other is approved 
final_filter <- group_T1w[group_T1w$bids_name %in% c("sub-CMH00000019_ses-01_run-02_T1w","sub-CMH00000039_ses-01_run-02_T1w", "sub-CMH00000056_ses-01_run-01_T1w", "sub-CMH00000270_ses-01_run-01_T1w", "sub-CMH00000287_ses-01_run-01_T1w", "sub-CMH00000435_ses-01_run-01_T1w" ), ]
final_filter

#### remove subsetted data frames ####
# remove the subsetted blacklist data frame using the anti_join function
filtered_T1 <- anti_join(group_T1w, blacklist_T1, by = "bids_name")
filtered_T1

# remove the subsetted flagged data frame using the anti_join function
extra_filtered_T1 <- anti_join(filtered_T1, flagged_T1, by = "bids_name") %>% select(bids_name, qi_1)
extra_filtered_T1

#remove the subsetted flagged data frame using the anti_join function
final_filtered_T1 <- anti_join(extra_filtered_T1, final_filter, by = "bids_name") %>% select(bids_name, qi_1)
final_filtered_T1

# rename bids_name column "participant" and string format to match brainage_final
# remove "_run-01"
final_filtered_T1$bids_name <- gsub("_run-01", "", final_filtered_T1$bids_name) 
# remove "_run-02"
final_filtered_T1$bids_name <- gsub("_run-02", "", final_filtered_T1$bids_name) 
# remove '_ses-01_T1w' string ending and rename 'bids_name' column 
final_filtered_T1 <- final_filtered_T1 %>%
  mutate(participant = str_replace(bids_name, "_ses-01_T1w", "")) %>%
  select(-bids_name) %>%
  relocate(participant, .before = qi_1)
final_filtered_T1

# save the filtered data frame as csv 
write.csv(final_filtered_T1, "/KIMEL/tigrlab/scratch/csun/brainage_proj/data/processed/mriqc_qi_1.csv", row.names = FALSE)

```


## Merge data frames
Create one data frame that includes BAG, centile scores, and ASD diagnoses
```{r Create one df w/ all variables}
# Append centile scores to enigma data frames
brainage_df <- merge(enigma_new,
                       centiles_final[c('participant', 'centile')], by ='participant', all.x = TRUE)
# merge by 'participant' column and append centile scores from centiles_final to enigma_new  

# Append ASD diagnoses
brainage_merged <- merge(
  brainage_df,enigma_ASD[c('participant', 'ASD_dx')], by ='participant', all.x = TRUE)

# Reorder columns
brainage_merged <- select(brainage_merged, participant, sex, ASD_dx, dx, total_dx, total_dx_group, age_band, Age, age_prediction, BAG, centile)
brainage_merged

# Create a "final" version of the merged data frame that excludes participants who are missing PSS/Non-PSS 'dx' info
brainage_final <- brainage_merged %>% 
  filter(!is.na(dx)) # only keep rows that are not missing 'dx'
brainage_final

# filter for NA for PSS and Non-PSS (~13)

# Append mriqc metrics qi_1 to brainage data frame
brainage_final <- merge(
  brainage_final, final_filtered_T1[c('participant', 'qi_1')], by ='participant', all.x = TRUE)
brainage_final
```

```{r Create new data frames separated by sex}

# females only
brainage_female <- filter(brainage_final, sex == "Female")

asd_female <- filter(enigma_ASD, sex == "Female")

# males only
brainage_male <- filter(brainage_final, sex == "Male")

asd_male <- filter(enigma_ASD, sex == "Male")

```

Questions related to sex differences:
1) Why is the predicted age significantly older for males?
2) Why is brain age gap significant larger (and more positive) for males?

## Save new data frames as csv
```{r Save the final merged brainage df}
write.csv(brainage_final, "/KIMEL/tigrlab/scratch/csun/brainage_proj/data/processed/brainage_final.csv", row.names = FALSE)
```

## Data visualization 

### What are the demographic variables we are working with?
1) Age - scan age
2) sex - sex assigned at birth
3) dx - diagnosis for PSS or Non-PSS
4) ASD_dx - indicates if participant has been diagnosed with ASD based on KSADS (only applicable to participants < age 18)
5) age_band - derived grouping (11-14, 15-17, 18-21, 21+) 
6) total_dx - numerical value representing the summed total number of disorders
7) total_dx_group - derived grouping (0-2, 3-4, 5+)

### What are the computed variables we are working with?
1) age_prediction - brain age predicted by ENIGMA
2) BAG - brain age gap calculated as the difference between brain age and scan age (age_prediction - Age)
3) centile - centile score predicted by BrainChart
4) qi_1 - QA metrics 

### Based on the above variables, what relationships can we visualize?
- Our main question for this project: how does brain age differ between individuals with and without psychosis spectrum symptoms (PSS)? 
- Secondary question: can differences in brain age be explained in part by sex, chronological age, multimorbidity, or ASD?
- Broader next steps: do our findings support the use of brain age and brain age gap as biological markers for the prognosis and/or prediction of mental illness?

```{r Check for background effects of scan quality }
# QI_1 vs predicted age 
ggplot(data = brainage_final, aes(x = age_prediction, y = qi_1)) +
       geom_point()
# There is one participant with extremely large QI_1 value
# check which participant using which.max
brainage_final[which.max(brainage_final$qi_1), ]
# participant 394 has QI_1 value 0.004482072

# what about participant 263 who has qi_1 of 0.01418? This is an even larger value than participant 394's. Participant 263 has already been removed i.e., not present in the final brainage data frames.

# remove participant 394 and visualize 
brainage_qi_1 <- filter(brainage_final, participant != "sub-CMH00000394")
brainage_qi_1

# qi_1 and age
ggplot(data = brainage_qi_1, aes(x = age_prediction, y = qi_1, color = Age)) +
  geom_jitter(width = .2)

# QI_1 vs centile scores
ggplot(data = brainage_qi_1, aes(x = centile, y = qi_1)) +
  geom_jitter(width = .2)

```

Proportions can be calculated using the mean() function: https://www.r-bloggers.com/2016/07/proportions-with-mean/
```{r proportions male vs female}
#### PSS ####
# What is the relative proportion of participants w/ PSS by sex?

# males
mean(brainage_male$dx == "PSS") # 0.414 or ~41%

# females
mean(brainage_female$dx == "PSS") # 0.56 or 56%

#### ASD ####
# What is the relative proportion of participants w/ ASD by sex?
# use enigma_ASD data frame
# males
mean(asd_male$ASD_dx == "Yes") # 0.267 or ~26.7%

# females
mean(asd_female$ASD_dx == "Yes") # 0.146 or ~14.6%


```


## Visualize the variation w/in and between variables
Resource for mapping variable values to colors: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/ 
```{r Set colors for sex + dx + ASD + qi_1}
# set colours for sex assigned at birth
female_colour <- "#FF0000" # red
male_colour <- "#007ED6" # blue

# alternative colours for sex
colour_f <- "#CC79A7" # pink
colour_m <- "#56B4E9" # light blue

# set colours for PSS and Non-PSS 
# colour_pss <- "#702A8C" # purple 
colour_pss <- "#6E298D" # CAMH purple
# colour_non_pss <- "#FFC154" # orange
colour_non_pss <- "#FFB652" # CAMH golden yellow

# set colours for ASD 
colour_ASD <- "#6B3E2E" # brown
  
# set colours for participants with non-zero 
colour_qi_1 <- "#D55E00" # red for QI1 values > 0
colour_qi_extreme <- "#AF4154" # maroon for QI1 values > 0.05e-04
# or use this alternative "old brick" colour: #96281B  
  
```


```{r BAG vs centile scores grouped by PSS grouping}

# plot BAG vs centile scores grouped by PSS and Non-PSS
ggplot(data = brainage_final, aes(x = BAG, y = centile, colour = dx, shape = dx)) +
       geom_point(aes(colour = dx)) + # set the colour based on PSS/Non-PSS
  geom_smooth(method=lm, linewidth = 0.5) +
  # Add linear regression line that includes 95% confidence region by default
 scale_y_continuous(limits = c(-0.2,1.2), expand = c(0.05,0)) +
  # use scale_y_continuous to set the range of the y-axis, expand to add some padding to ensure no values are cut-off
  theme_classic() +
  facet_wrap(~age_band) +
  scale_color_manual(values = c("PSS" = colour_pss, "Non-PSS" = colour_non_pss)) + # set colours for PSS grouping
   xlab("ENIGMA Derived Brain Age Gap") +
  ylab("BrainChart Centile Score") +
  labs(colour = "") + 
   guides(colour = guide_legend(data = brainage_final, override.aes = list(shape = c(16,17)))) 

```


```{r BAG vs centile scores grouped by sex and age_band}
# set colours for sex assigned at birth
female_colour <- "#FF0000" # Red
male_colour <- "#007ED6" # Blue
# plot BAG vs centile scores grouped by sex
ggplot(data = brainage_final, aes(x = BAG, y = centile, colour = sex)) +
       geom_point(shape = 1) + # Use hollow circles
  geom_smooth(method=lm, linewidth = 0.5) +
  # Add linear regression line that includes 95% confidence region by default
  scale_y_continuous(limits = c(-0.15,1.15), expand = c(0,0)) +
  # use scale_y_continuous to set the range of the y-axis
  theme_classic() +
  facet_wrap(~age_band) +
   scale_colour_manual(values = c("Male" = colour_m, "Female" = colour_f)) + # set colours for sex
   xlab("ENIGMA Derived Brain Age Gap") +
  ylab("BrainChart Centile Score") +
  labs(colour = "Sex Assigned at Birth")
```

```{r BrainAGE vs centile scores grouped by age band and sex and dx}
# plot BAG vs centile scores grouped by age band, sex, and PSS
ggplot(data = brainage_final, aes(x = BAG, y = centile, colour = sex, shape = dx)) +
       geom_point(aes(colour = sex, shape = dx)) + # try setting the shape to represent PSS and Non-PSS
  geom_smooth(method=lm, se = FALSE, linewidth = 0.5) + # remove standard error shading using se = FALSE
  # Add linear regression line that includes 95% confidence region by default
  scale_y_continuous(limits = c(-0.25,1.2), expand = c(0,0)) +
  # use scale_y_continuous to set the range of the y-axis
  theme_classic() +
  facet_wrap(~age_band) +
   scale_colour_manual(values = c("Male" = colour_m, "Female" = colour_f)) + # set colours for sex
   xlab("ENIGMA Derived BrainAGE") +
  ylab("BrainChart Centile Score") +
  labs(colour = "Sex Assigned at Birth")
```

```{r BrainAGE vs centile scores grouped by sex and PSS Non-PSS}

# plot BrainAGE vs centile scores grouped by PSS and facet wrap by sex
ggplot(data = brainage_final, aes(x = BAG, y = centile, colour = dx, shape = dx)) +
       geom_point(aes(colour = dx, shape = dx)) + # try setting the shape to represent PSS and Non-PSS
  geom_smooth(method=lm, linewidth = 0.5) + # remove standard error shading using se = FALSE
  # Add linear regression line that includes 95% confidence region by default
  scale_y_continuous(limits = c(-0.35,1.2), expand = c(0,0)) +
  # use scale_y_continuous to set the range of the y-axis
  theme_classic() +
  facet_wrap(~sex) +
   scale_colour_manual(values = c("PSS" = colour_pss, "Non-PSS" = colour_non_pss)) + # set colours for PSS/Non-PSS
   xlab("ENIGMA Derived BrainAGE") +
  ylab("BrainChart Centile Score") +
  labs(colour = "", shape = "")
```


```{r BAG vs centile scores grouped by sex and total_dx}
# plot BAG vs centile scores grouped by sex
ggplot(data = brainage_final, aes(x = BAG, y = centile, colour = sex)) +
       geom_point(shape = 1) + # Use hollow circles
  geom_smooth(method=lm) +
  # Add linear regression line that includes 95% confidence region by default
  scale_y_continuous(limits = c(0,1), expand = c(0,0)) +
  # use scale_y_continuous to set the range of the y-axis
  theme_classic() +
  facet_wrap(~total_dx) + 
  scale_colour_manual(values = c("Male" = colour_m, "Female" = colour_f)) +
  xlab("ENIGMA Derived Brain Age Gap") +
  ylab("BrainChart Centile Score")

# Source: http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/
```

```{r Interactions b/w PSS and ASD}
# boxplot 
ggplot(data = subset(brainage_final, !is.na(ASD_dx)), aes(x = dx, y = centile, colour = ASD_dx)) +
       geom_boxplot() +
  theme_classic() +
 # facet_wrap(~ASD_dx) + 
  ylab("BrainChart Centile Score") +
  xlab("PSS grouping") +
  labs(title = "Centile Scores by ASD and PSS Diagnoses")

```


```{r Correlation test for centile vs BAG}
cor.test(brainage_final$centile, brainage_final$BAG, method = "pearson")

# print the actual p-value 
cor.test(brainage_final$centile, brainage_final$BAG, method = "pearson")$p.value
```
The scatter plot suggests a negative correlation between centile scores and brain age gap for both PSS and Non-PSS groups. However, the BAG range for PSS participants appears to broader (longer line). This relationship is confirmed by Pearson's correlation test (r = -0.63, p < 0.0001).

```{r centile vs age-band grouped by dx and sex}
#  centile scores PSS and Non-PSS by age bands
brainage_final %>%  
  ggplot(aes(x = age_band, y = centile, colour = dx)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic() +
  facet_wrap(~sex) +
  scale_color_manual(values = c("PSS" = colour_pss, "Non-PSS" = colour_non_pss)) + # set colours for PSS grouping
   xlab("ENIGMA Derived Brain Age Gap") +
  ylab("BrainChart Centile Score")
  
# 3 way ANOVA 
```

```{r BAG vs age_band grouped by sex and dx}

# BAG PSS and Non-PSS for by age categories
brainage_final %>% drop_na(ASD_dx) %>% 
  filter(age_band %in% c("15-17", "11-14")) %>%
  # filter out the two older age categories b/c KSADS only done for participants age < 18
  ggplot(aes(x = age_band, y = BAG, colour = sex)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic() +
  facet_wrap(~dx) +
  scale_colour_manual(values = c("Male" = colour_m, "Female" = colour_f)) +
   ylab("ENIGMA Derived Brain Age Gap") +
  xlab("Age Band")
```
```{r BAG vs age_band grouped by multimorbidity}
# BAG PSS and Non-PSS for by total disorder grouping
brainage_final %>%  
  ggplot(aes(x = age_band, y = BAG, colour = total_dx_group)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic() +
   ylab("ENIGMA Derived Brain Age Gap") +
  xlab("Age Band") +
  labs(colour = "Morbidity Group")
```


```{r centile scores vs age_band grouped by multimorbidity}
# Centile PSS and Non-PSS for by total disorder grouping
brainage_final %>%  
  ggplot(aes(x = age_band, y = centile, colour = total_dx_group)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic() +
  ylab("BrainChart Centile Score") +
  xlab("Age Band") +
  labs(colour = "Morbidity Group")
```


```{r BAG vs age_band grouped by multimorbidity and sex}
# these groupings may not be valid b/c there are some age bands with only one participant

# BAG PSS and Non-PSS for by total disorder grouping
brainage_final %>%  
  ggplot(aes(x = age_band, y = BAG, colour = total_dx_group)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic() + 
  facet_wrap(~sex) + 
    ylab("ENIGMA Derived Brain Age Gap") +
  xlab("Age Band") +
  labs(colour = "Morbidity Group")
```


Q: why is the average centile score lower for individuals with PSS?
Q: why is the average BAG greater for individuals with ASD?

```{r Variation with ASD}
# plot BAG among participants with and without ASD
ggplot(data = subset(brainage_final, !is.na(ASD_dx)), aes(x = ASD_dx, y = BAG)) +
       geom_boxplot() +
  theme_classic()

# centile scores
ggplot(data = subset(brainage_final, !is.na(ASD_dx)), aes(x = ASD_dx, y = centile, colour = age_band)) +
       geom_boxplot() +
  theme_classic() 

brainage_final %>% drop_na(ASD_dx) %>% 
  filter(age_band %in% c("15-17", "11-14")) %>%
  # filter out the two older age categories b/c KSADS only done for participants age < 18
  ggplot(aes(x = age_band, y = centile, colour = ASD_dx)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9)) +
  # add geom_point() to overlay the points
       theme_classic()  
# BAG
brainage_final %>% drop_na(ASD_dx) %>% 
  filter(age_band %in% c("15-17", "11-14")) %>%
  ggplot(aes(x = age_band, y = BAG, colour = ASD_dx)) +
   geom_boxplot(position=position_dodge(width=0.9)) +
  geom_point(position=position_jitterdodge(dodge.width=0.9))

# Compare BAG and centile scores between participants with and without ASD
ggplot(data = subset(brainage_final, !is.na(ASD_dx)), aes(x = BAG, y = centile, colour = ASD_dx)) +
       geom_point(shape = 1) + # Use hollow circles
  geom_smooth(method=lm) +
  # Add linear regression line that includes 95% confidence region by default
  scale_y_continuous(limits = c(0,1), expand = c(0,0)) +
  # use scale_y_continuous to set the range of the y-axis
  theme_classic() 
# ADD FACET WRAP HERE

# look at PSS and ASD
```


```{r Explore variation with age bands}
# use the reorder() function within aes to reorder the x variable based on the median value of the y variable

# Age prediction 
ggplot(data = brainage_final, aes(x = reorder(age_band, age_prediction, FUN = median), y = age_prediction)) + 
  geom_boxplot()

# Brain age gap
ggplot(data = brainage_final, aes(x = reorder(age_band, BAG, FUN = median), y = BAG)) + 
  geom_boxplot()

# Centile scores
ggplot(data = brainage_final, aes(x = reorder(age_band, centile, FUN = median), y = centile)) + 
  geom_boxplot()

```
The box plot shows that the 15-17 age band has the largest range of centile scores.

```{r Explore variation with total disorder groups}
# Age prediction
ggplot(data = brainage_final, aes(x = reorder(total_dx_group, age_prediction, FUN = median), y = age_prediction)) + 
  geom_boxplot()

# Brain age gap
ggplot(data = brainage_final, aes(x = reorder(total_dx_group, BAG, FUN = median), y = BAG)) + 
  geom_boxplot()

# Centile score
ggplot(data = brainage_final, aes(x = reorder(total_dx_group, centile, FUN = median), y = centile)) + 
  geom_boxplot()

```

## Pairwise comparisons by sex
```{r t-test for total_dx and sex}
# need to set categorical variables as two-level factor before conducting t tests
# sex: Male, Female
brainage_final$sex <- as.factor(brainage_final$sex)
# ASD: Yes, No
enigma_ASD$ASD_dx <- as.factor(enigma_ASD$ASD_dx)
# dx: PSS, Non-PSS
brainage_final$dx <- as.factor(brainage_final$dx)


# sex and total_dx
t.test(total_dx ~ sex, alternative = "two.sided", data = brainage_final)

# for more t-tests see the "Comparisons" section of "04_explore_analyses_and_stats.Rmd"
```

```{r chi-square tests to compare 2 categorical vars}
# sex and dx
table_dx <- table(brainage_final$dx, brainage_final$sex)
result_dx <- chisq.test(table_dx)
result_dx

# sex and ASD_dx
table_asd <- table(enigma_ASD$ASD_dx, enigma_ASD$sex)
result_asd <- chisq.test(table_asd)
result_asd

# dx and age_band
table_age_band <- table(brainage_final$dx, brainage_final$age_band)
result_age_band <- chisq.test(table_age_band)
result_age_band

# sex and age_band
table_ageb_sex <- table(brainage_final$sex, brainage_final$age_band)
result_ageb_sex <- chisq.test(table_ageb_sex)
result_ageb_sex
# 
```
Results of Pearson's Chi-squared test 
- for sex and dx: X-squared = 4.6202, df = 1, p-value = 0.0316
- for sex and ASD_dx: X-squared = 2.318, df = 1, p-value = 0.1279
- for dx and age_band: X-squared = 1.3204, df = 3, p-value = 0.7243
- for sex and age_band: X-squared = 10.426, df = 3, p-value = 0.01527


## Linear models
```{r t tests for ASD}

# ASD: BAG vs ASD_dx
# ASD: centile vs ASD_dx
# see "Comparisons" section of "04_explore_analyses_and_stats.Rmd" for ASD t tests
```


```{r ANOVA linear models for centile scores}

# centile ANOVA

lm_centile <- lm(centile ~ sex * dx * age_band, data = brainage_final)

aov(lm_centile) # perform ANOVA on the linear model
plot(aov(lm_centile)) # verify assumptions (normal distribution, homoscedasticity, additivity)
summary(aov(lm_centile)) # identify strength of interactions

# age
lm_centile_age <- lm(centile ~ sex * dx * Age, data = brainage_final)
aov(lm_centile_age)
summary(aov(lm_centile_age))

# total_dx
lm_centile_total_dx <- lm(centile ~ sex * dx * total_dx, data = brainage_final)
aov(lm_centile_total_dx)
summary(aov(lm_centile_total_dx))

# total_dx_group
lm_centile_total_dx_grp <- lm(centile ~ sex * dx * total_dx_group, data = brainage_final)
aov(lm_centile_total_dx_grp)
summary(aov(lm_centile_total_dx_grp))
```

```{r ANOVA linear models for BAG}
# BAG ANOVA

## age bands
lm_BAG <- lm(BAG ~ sex * dx * age_band, data = brainage_final)
aov(lm_BAG)
summary(aov(lm_BAG))
# sex: p = 0.000111
# dx: p = 0.079625
# age_band: p = 0.149424

## scan age
lm_age <- lm(BAG ~ sex * dx * Age, data = brainage_final)
aov(lm_age)
summary(aov(lm_age))

## total dx
lm_total_dx <- lm(BAG ~ sex * dx * total_dx, data = brainage_final)
aov(lm_total_dx)
summary(aov(lm_total_dx))

## total dx group
lm_total_dx_grp <- lm(BAG ~ dx * sex * total_dx_group, data = brainage_final)
aov(lm_total_dx_grp)
summary(aov(lm_total_dx_grp))
```

```{r chi-square test comparing proportion of male and female sex by age band}

#### Method 1: identifying differences across ALL age bands ####
# Note: this is not the most accurate method b/c it does not account for differences in sample size (e.g., more females than males)
# convert sex, age_band to factors
# age_band: 11-14, 15-17, 18-21, 21+
brainage_final$age_band <- as.factor(brainage_final$age_band)
# sex: Female, Male
brainage_final$sex <- as.factor(brainage_final$sex)
brainage_final

# Create a contingency table of 'sex' and 'age_band'
contingency_table <- table(brainage_final$sex, brainage_final$age_band)
contingency_table # there are more than 5 participants per group so chi-square test assumptions are met

# perform the chi-square test
result <- chisq.test(contingency_table)

# print the result
print(result)

#### Method 2: ANOVA for proportion sex and age band ####

# calculate the total number of participants in each age band
total_by_age_band <- aggregate(list(TotalParticipants = brainage_final$sex), by = list(brainage_final$age_band), length)

# calculate the number of male participants in each age band
male_by_age_band <- aggregate(list(MaleParticipants = brainage_final$sex[brainage_final$sex == "Male"]), by = list(brainage_final$age_band[brainage_final$sex == "Male"]), length)

# calculate the proportion of males
prop_by_age_band_male <- merge(total_by_age_band, male_by_age_band, by = "Group.1", all.x = TRUE)
prop_by_age_band_male$ProportionMale <- prop_by_age_band_male$MaleParticipants / prop_by_age_band_male$TotalParticipants

# Print the results
print(prop_by_age_band_male)

# load csv with proportions for each sex calculated based on total number of participants of that sex - this is different from the proportions calculated based on total # for each age band
chi_sq_df <- read_csv("/KIMEL/tigrlab/scratch/csun/brainage_proj/data/raw/chi_square.csv")
chi_sq_df

# Perform ANOVA to
# convert sex, age_band to factors
# age_band: 11-14, 15-17, 18-21, 21+
chi_sq_df$age_band <- as.factor(chi_sq_df$age_band)
# sex: Female, Male
chi_sq_df$sex <- as.factor(chi_sq_df$sex)
chi_sq_df

# perform ANOVA test
result_prop_by_sex <- aov(prop ~ age_band * sex, data = chi_sq_df)
print(result_prop_by_sex)
summary(result_prop_by_sex)

```
Based on Pearson's Chi-squared test, there is a significant difference in the proportion of male vs female participants across age bands (p=0.01527, X-squared = 10.426)

```{r chi square test PSS grouping vs age_band}
# convert sex, age_band to factors
# age_band: 11-14, 15-17, 18-21, 21+
brainage_final$age_band <- as.factor(brainage_final$age_band)
# dx (aka PSS grouping): PSS, Non-PSS
brainage_final$dx <- as.factor(brainage_final$dx)
brainage_final

# Create a contingency table of 'sex' and 'age_band'
contingency_table_dx <- table(brainage_final$dx, brainage_final$age_band)
contingency_table_dx # there are more than 5 participants per group so chi-square test assumptions are met

# perform the chi-square test
result2 <- chisq.test(contingency_table_dx)

# print the result
print(result2)

```
There are no significant differences (p=0.7243) in the proportion of participants with and without PSS across the age bands.

## Extra analyses
```{r Additional analyses for age}
mean(brainage_final$Age) # 18.47
mean(brainage_final$age_prediction) # 19.73
```


```{r Additional sex based analyses - female}
#### Correlation between centile score and BrainAGE ####
cor.test(brainage_female$centile, brainage_female$BAG, method = "pearson")
# r = -0.7026, p<0.001

# print the actual p-value 
cor.test(brainage_female$centile, brainage_female$BAG, method = "pearson")$p.value

#### t-test comparing centile score vs PSS grouping ####
# need to set categorical variables as two-level factor before conducting t tests
# sex: Female
brainage_female$sex <- as.factor(brainage_female$sex)
# dx: PSS, Non-PSS
brainage_female$dx <- as.factor(brainage_female$dx)

# centile and PSS/Non-PSS
t.test(centile ~ dx, alternative = "two.sided", data = brainage_female)
# PSS=0.4301038, Non-PSS=0.5398383, p-value=0.005686
```
The correlation between BrainChart centile score and ENIGMA BrainAGE for females is -0.7026 (p < 2.2e-16) based on Pearson's correlation test.

Females with PSS have significantly lower BrainChart centile scores (mu=0.430, p=0.006) compared to females without PSS (mu=0.540).

```{r Additional sex based analyses - males}
#### Correlation between centile score and BrainAGE ####
cor.test(brainage_male$centile, brainage_male$BAG, method = "pearson")
# r = -0.619, p=1.641e-10

# print the actual p-value 
cor.test(brainage_male$centile, brainage_male$BAG, method = "pearson")$p.value

#### t-test comparing centile score vs PSS grouping ####
# need to set categorical variables as two-level factor before conducting t tests
# sex: male
brainage_male$sex <- as.factor(brainage_male$sex)
# dx: PSS, Non-PSS
brainage_male$dx <- as.factor(brainage_male$dx)

# centile and PSS/Non-PSS
t.test(centile ~ dx, alternative = "two.sided", data = brainage_male)
# PSS=0.564, Non-PSS=0.499, p-value=0.313

```
The correlation between BrainChart centile score and ENIGMA BrainAGE for males is -0.619 (p=1.641e-10) based on Pearson's correlation test.

Males with PSS had lower centile scores on average (mu=0.499) than those without PSS (mu=0.564); however, this difference was not significant.


```{r testing Citrix}
# number of individuals with and without PSS

# install.packages("psych")
library(psych)
describe(brainage_final)
# display summary statistics table for female participants 
describe(brainage_female)
# display summary statistics table for male participants 
describe(brainage_male)

```


```{r testing Citrix 2}
print("hello")
# hi it's me
# hello from the other side
# testing 123 now it is lagging
# still here
# saying hello from the corner of Gerstein 2nd floor
# still here - met the caretaker cleaning the women's bathroom, seemed friendly
# testing 
# 123