---
title: "Principal Component Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Resources
How to create a correlation matrix - http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2 

## Set up working environment
```{r For local users only - install packages}
install.packages("ggcorrplot")
# ggcorrplot includes functions to reorder the correlation matrix and display significant level on the plot; computes matrix of correlation p-values

install.packages("stats")
# includes functions for principal component analysis

install.packages("devtools") 
install_github("kassambara/factoextra")

```


```{r Set up - Load libraries}
library(tidyverse)
library(ggcorrplot)
library(dplyr)
library(stats)
library(devtools)
# library("factoextra") # unable to load 
```

```{r Set up - Load and tidy data}
# Load brainage_final
brainage_final <- read_csv("/KIMEL/tigrlab/scratch/csun/brainage_proj/data/processed/brainage_final.csv")
# brainage_final.csv contains 12 columns (participant ID + 11 variables of interest): sex, ASD diagnoses, PSS dx, total dx, total dx group, age band, scan age, ENIGMA age prediction, ENIGMA BAG, BrainChart centile score, qi_1

# Subset brainage_final to keep only numeric columns and remove NAs
brainage_num <- brainage_final %>% 
  select(where(is.numeric)) %>% # subset by numeric columns
  na.omit(brainage_final) # omit all rows with NA values in any column
 # filter(!is.na(qi_1)) # another way to remove row(s) with NA values by removing NA values from a specific column (in this case qi_1 is the only column with NAs)
# brainage_num

# we are removing non-numeric columns because the correlation matrix is intended for numeric data 
# resource: https://www.statology.org/dplyr-select-numeric-columns/
```


```{r Set up - subset data to 5 variables}
# Subset brainage_num further to exclude the qi_1 column because its correlation values are closed to 0
brainage_5 <- brainage_num %>%
  select(-qi_1) # remove the qi_1 column
brainage_5

```
```{r Set up - subset data to only age related variable}
brainage <- brainage_final %>% 
  select(where(is.numeric)) %>% # subset by numeric columns
  na.omit(brainage_final) %>%
  select(-qi_1) %>% # remove the qi_1 column
  select(-total_dx) # remove the total_dx column

brainage
```


## Correlation Matrix
```{r Compute a correlation matrix for 6 variables}

# Compute a correlation matrix of p-values
corr <- round(cor(brainage_num), 1)

# view first 6 rows of the correlation matrix
head(corr[, 1:6])
```

```{r Compute a correlation matrix for 5 variables}
# Compute a correlation matrix of p-values
corr2 <- round(cor(brainage_5), 1)

# view first 5 rows of the correlation matrix
head(corr2[, 1:5])
```
```{r Compute a correlation matrix for age related variables}
corr3 <- round(cor(brainage), 1)

# view first 5 rows of the correlation matrix
head(corr3[, 1:4])
```


```{r Compute a matrix of correlation p-values for 6 variables}
# use the cor_pmat() function from corrplot pkg
p_mat <- cor_pmat(brainage_num)

# display the full p-value matrix
head(p_mat[, 1:6])
```

```{r Compute a matrix of correlation p-values for 5 variables}
# use the cor_pmat() function from corrplot pkg
p_mat2 <- cor_pmat(brainage_5)

# display the full p-value matrix
head(p_mat2[, 1:5])
```
```{r Compute a matrix of correlation p-values for age variables}
# use the cor_pmat() function from corrplot pkg
p_mat3 <- cor_pmat(brainage)

# display the full p-value matrix
head(p_mat3[, 1:4])
```

## Correlation matrix visualization 
```{r Visualize the correlation matrix for 6 variables}
# use the default "square" method
ggcorrplot(corr)

# reorder the correlation matrix using hierarchical clustering
ggcorrplot(corr, hc.order = TRUE, outline.col = "white")

# add correlation coefficients with argument lab = TRUE
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE)
```
Observations from correlation matrix w/ 6 variables:
- qi_1 has very low correlation with all variables
- BAG and predicted age are highly correlated (this is expected because BAG was derived from the predicted age)

```{r Visualize the correlation matrix for 5 variables}
# use the default "square" method
ggcorrplot(corr2)

# reorder the correlation matrix using hierarchical clustering
ggcorrplot(corr2, hc.order = TRUE, outline.col = "white")

# add correlation coefficients with argument lab = TRUE
ggcorrplot(corr2, hc.order = TRUE, type = "lower", lab = TRUE)
```

```{r Visualize the correlation matrix for age variables}
# use the default "square" method
ggcorrplot(corr3)

# reorder the correlation matrix using hierarchical clustering
ggcorrplot(corr3, hc.order = TRUE, outline.col = "white")

# add correlation coefficients with argument lab = TRUE
ggcorrplot(corr3, hc.order = TRUE, type = "lower", lab = TRUE)
```

## Principal Component Analysis (PCA)
```{r PCA eligibility}
# unsupervised learning - explore data without labels

# check PCA eligibility 
# are the variables highly correlated? (linearly dependent or not)
# correlation matrix to check for degree of correlation, assuming high degree of correlation if r > 0.3 or r < -0.3

# average correlation among the variables
mean(corr) # mean for correlation of all 6 variables is 0.18889 
mean(corr2) # mean for correlation of 5 variables is 0.248 -> indicates variables are not highly correlated - does this mean the data are not eligible?
mean(corr3)

```
NOTE: the PCA below may not be valid, need to double check eligibility requirements above
```{r Conduct PCA}
pca <- princomp(brainage_num) 
pca2 <- princomp(brainage_5)
pca3 <- princomp(brainage)
```

## Evaluating PCA
- remember: the number of components is always equal to the number of variables 
Two ways to evaluate PCA
1. Check whether principal components (PCs) capture the essence of the original variable
2. Check whether PCs are independent
- Resource: https://youtu.be/jFN4qkSOd4I

```{r Inspect PC loadings}
# inspect contents of pca
pca$loadings # 6 components bc 6 variables in original df
pca2$loadings # 5 components
pca3$loadings # 4 components

# focus on the first two components
pca$loadings[,1:2]
# first two components of 5 PC 
format(pca2$loadings[,1:2], scientific = FALSE)
# first two components of 4 PC
format(pca3$loadings[,1:3], scientific = FALSE)
```

```{r Check for independence among principal components}

# PCs are considered independent if the correlation among PCs is approximately 0
pc <- pca$scores # first create an object with scores
pc2 <- pca2$scores
pc3 <- pca3$scores
# compare correlation scores for PCs
cor(pc) 
# correlation for PCs is almost zero EXCEPT comp 6 -due to qi_1?
cor(pc2) # does not seem like component 5 is independent
cor(pc3)
```
```{r Visualization of the principal components}
#### Scree plot ####

# calculate total variance explained by each principal component
var_explained = pca$sdev^2 / sum(pca$sdev^2)
var_explained

qplot(c(1:6), var_explained) +
  geom_line() +
  # geom_bar() +
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0,1)

# resource: https://www.statology.org/scree-plot-r/

```

