---
title: "T1w Image Quality Metrics Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r Set up - Load libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
```

## Load data

We will be working with four T1w image quality metrics (IQM) calculated from MRIQC for structural MRI data:
1. CJV = coefficient of joint variation (cjv) - lower better
2. SNRD = Dietrich's signal-to-noise ratio, total (snrd_total) - higher better
3. EFC = entropy focus criterion (efc) - lower better
4. FBER = foreground-to-background energy ratio (fber) - higher better
5. FWHM = full-width, half-maximum (fwhm_avg)

*Reference TAY Cohort Project 3 Protocol Paper for more information on how the data was obtained
*See Supplemental Figure 5 from the protocol paper "T1w quality Metrics (from MRIQC)" which includes signal to noise, FWHM, FBER, CJV

```{r Set up - Load csv and tsv files as data frames}
# Set up code modified from 06_mri_qi1_effects.Rmd and 05_merged_stats.Rmd

# brainage_final includes participants' demographic info and brain age outputs from ENIGMA and BrainChart
brainage_final <- read_csv("/KIMEL/tigrlab/scratch/csun/brainage_proj/data/processed/brainage_final.csv")
# brainage_final

# update participant 433's qi_1 value
brainage_final <- brainage_final %>% mutate(qi_1 = coalesce(qi_1, 0)) # use the coalesce() function to replace NA with zero in a specific numeric column
brainage_final

# group_T1w contains T1w image quality metrics (IQMs)
group_T1w <- read_delim("/external/rprshnas01/tigrlab/archive/data/TAY/pipelines/versioned_outputs/baseline_only/mriqc/group_T1w.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
group_T1w

# participant 433's IQMs
sub_433_T1w <- read_delim("/KIMEL/tigrlab/scratch/jwong/mriqc_sub433/output/group_T1w.tsv", delim = "\t", escape_double = FALSE, trim_ws = TRUE)

# append participant 433's T1w IQMs to end of group_T1w df
group_T1w_full <- rbind(group_T1w, sub_433_T1w)

```
Resource on appending data to a data frame: https://www.statology.org/r-append-to-data-frame/

## Tidy data in preparation for PCA
Merge multiple runs based on same criteria applied to qi_1 analyses (see "Incorporate QA metrics" in 05_merged_stats.Rmd)

Assumptions for removing and merging qi_1 values for participants with multiple runs, double checked with Erin and Jimmy
- If one of the runs is a blacklisted scan, remove it and keep the remaining scan
- If one run is flagged while the other is approved, remove the flagged scan and keep the approved
- If both runs are flagged, keep the run with the smaller qi_1 value
- If both runs have qi_1 value of 0, merge and keep the qi_1 value at 0

```{r Merge participants with multiple runs}
# Merging multiple runs by qi_1 data
#### remove qi_1 values for blacklisted scans ####
blacklist <- c("sub-CMH00000218_ses-01_run-01_T1w", "sub-CMH00000224_ses-01_run-02_T1w", "sub-CMH00000283_ses-01_run-01_T1w", "sub-CMH00000421_ses-01_run-01_T1w", "sub-CMH00000466_ses-01_run-01_T1w")
class(blacklist)

# note we are using group_T1w_full which includes appended participant 433
blacklist_T1 <- group_T1w_full[group_T1w_full$bids_name %in% c("sub-CMH00000218_ses-01_run-01_T1w", "sub-CMH00000224_ses-01_run-02_T1w", "sub-CMH00000283_ses-01_run-01_T1w", "sub-CMH00000421_ses-01_run-01_T1w", "sub-CMH00000466_ses-01_run-01_T1w"), ]
blacklist_T1

#### remove qi_1 values for flagged runs ####
# includes flagged runs that are paired with an approved run and flagged runs with larger qi_1 
flagged_T1 <- group_T1w_full[group_T1w_full$bids_name %in% c("sub-CMH00000062_ses-01_run-02_T1w", "sub-CMH00000150_ses-01_run-01_T1w", "sub-CMH00000192_ses-01_run-01_T1w", "sub-CMH00000263_ses-01_run-01_T1w", "sub-CMH00000360_ses-01_run-01_T1w", "sub-CMH00000361_ses-01_run-01_T1w", "sub-CMH00000369_ses-01_run-01_T1w"), ]
flagged_T1

#### runs that both have qi_1 values of 0 ####
# delete one of the rows, choose the one that is flagged if the other is approved 
final_filter <- group_T1w_full[group_T1w_full$bids_name %in% c("sub-CMH00000019_ses-01_run-02_T1w","sub-CMH00000039_ses-01_run-02_T1w", "sub-CMH00000056_ses-01_run-01_T1w", "sub-CMH00000270_ses-01_run-01_T1w", "sub-CMH00000287_ses-01_run-01_T1w", "sub-CMH00000435_ses-01_run-01_T1w" ), ]
final_filter

#### remove subsetted data frames ####
# remove the subsetted blacklist data frame using the anti_join function
filtered_T1 <- anti_join(group_T1w_full, blacklist_T1, by = "bids_name")
filtered_T1

# remove the subsetted flagged data frame using the anti_join function
extra_filtered_T1 <- anti_join(filtered_T1, flagged_T1, by = "bids_name") %>% select(bids_name, qi_1, cjv, cnr, efc, fber, fwhm_avg, snrd_total)
extra_filtered_T1

#remove the subsetted flagged data frame using the anti_join function
final_filtered_T1 <- anti_join(extra_filtered_T1, final_filter, by = "bids_name") %>% select(bids_name, qi_1, cjv, cnr, efc, fber, fwhm_avg, snrd_total)
final_filtered_T1

# rename bids_name column "participant" and string format to match brainage_final
# remove "_run-01"
final_filtered_T1$bids_name <- gsub("_run-01", "", final_filtered_T1$bids_name) 
# remove "_run-02"
final_filtered_T1$bids_name <- gsub("_run-02", "", final_filtered_T1$bids_name) 
# remove '_ses-01_T1w' string ending and rename 'bids_name' column 
final_filtered_T1 <- final_filtered_T1 %>%
  mutate(participant = str_replace(bids_name, "_ses-01_T1w", "")) %>%
  select(-bids_name) %>%
  relocate(participant, .before = qi_1)
final_filtered_T1

# save the filtered data frame as csv 
write.csv(final_filtered_T1, "/KIMEL/tigrlab/scratch/csun/brainage_proj/data/processed/T1w_quality_metrics.csv", row.names = FALSE)

```



```{r Merge IQM df and brain age df}

# Append MRIQC quality metrics df to brain age data frame
brainage_final_iqms <- merge(
  brainage_final, final_filtered_T1[c('participant', 'cjv', 'cnr', 'efc', 'fber', 'fwhm_avg', 'snrd_total')], by ='participant', all.x = TRUE)
brainage_final_iqms

```

The new brainage_final_iqms data frame includes all demographic data, brain age outputs from ENIGMA and BrainChart, and MRIQC image quality metrics.

1. CJV = coefficient of joint variation (cjv)
2. SNRD = Dietrich's signal-to-noise ratio, total (snrd_total)
3. EFC = entropy focus criterion (efc)
4. FBER = foreground-to-background energy ratio (fber)
5. FWHM = full-width, half-maximum (fwhm_avg)

## Data exploration 
```{r Visualize distribution of each IQM}
# check if each IQM is normally distributed using histograms

# coefficient of joint variation
hist(final_filtered_T1$cjv)

# Dietrich's signal-to-noise ratio, total
hist(final_filtered_T1$snrd_total)

# entropy focus criterion
hist(final_filtered_T1$efc)

# foreground-to-background energy ratio
hist(final_filtered_T1$fber)

# full-width, half maximum
hist(final_filtered_T1$fwhm_avg)
```

## Correlation Matrix
Compute a correlation matrix for PCA based on code included the "Correlation Matrix" section in 07_corr_and_pca.Rmd

```{r Calculating correlation matrix}
# first check for NAs
which(is.na(final_filtered_T1))
sum(is.na(final_filtered_T1)) # there are no NAs in this df

# subset final_filtered_T1 to only include numeric columns i.e., remove 'participant' column
T1w_iqms_num <- final_filtered_T1 %>% 
  select(where(is.numeric)) %>%
  select(-qi_1)

# Compute a matrix of Pearson correlation coefficients
corr <- round(cor(T1w_iqms_num), 3) # round to 3 decimal places
corr
# view first 6 rows of the correlation matrix
head(corr[, 1:6])

```
```{r Compute a matrix of correlation p-values}
p_mat <- cor_pmat(T1w_iqms_num)
p_mat

```

## Correlation Matrix Visualization
```{r Visualize the correlation matrix of correlation coefficients}
# use the default "square" method
ggcorrplot(corr)

# reorder the correlation matrix using hierarchical clustering
ggcorrplot(corr, hc.order = TRUE, outline.col = "white")

# add correlation coefficients with argument lab = TRUE
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE)
```

Observations from correlation matrix of 6 image quality metrics:
- IQMs where higher values are better (indicate higher quality scans) are positively correlated as expected: snrd_total and fber
- interesting relationship between efc and cjv -> expect positive correlation but observe negative correlation (?)
- strong negative correlation between cnr and cjv, this is consistent with the fact that higher values of cnr are better, and lower values of cjv are better

## Principal Component Analysis
```{r PCA eligibility}
# unsupervised learning - explore data without labels

# check PCA eligibility 
# are the variables highly correlated? (linearly dependent or not)
# correlation matrix to check for degree of correlation, assuming high degree of correlation if r > 0.3 or r < -0.3

# average correlation among the variables
mean(corr) # mean for correlation of all variables is 0.2173889
```
NOTE: the PCA below may not be valid, need to double check eligibility requirements above

```{r Conduct PCA}
pca <- princomp(T1w_iqms_num) 
```

## Evaluating PCA

- remember: the number of components is always equal to the number of variables 
Two ways to evaluate PCA
1. Check whether principal components (PCs) capture the essence of the original variable
2. Check whether PCs are independent
- Resource: https://youtu.be/jFN4qkSOd4I

```{r Inspect PCA loadings}
# inspect contents of pca
pca$loadings
pca$loadings[,1:4] # focus on the first 4 components
```

```{r Check for independence among principal components}
# PCs are considered independent if the correlation among PCs is approximately 0
pc <- pca$scores # first create an object with scores

# compare correlation scores for PCs
cor(pc) 

```

```{r Visualization of the principal component}
#### Scree plot ####

# calculate total variance explained by each principal component
var_explained = pca$sdev^2 / sum(pca$sdev^2)
var_explained

qplot(c(1:6), var_explained) +
  geom_line() +
  # geom_bar() +
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Scree Plot") +
  ylim(0,1)

```
Based on the screen plot, it looks like only the first principal component (PC1) is responsible for explaining the major of the variance
