---
title: "T1w Image Quality Metrics Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r Set up - Load libraries}
library(tidyverse)
library(dplyr)
library(ggplot)
```

## Load data

We will be working with four T1w image quality metrics (IQM) calculated from MRIQC for structural MRI data:
1. CJV = coefficient of joint variation (cjv)
2. SNRD = Dietrich's signal-to-noise ratio, total (snrd_total)
3. EFC = entropy focus criterion (efc)
4. FBER = foreground-to-background energy ratio (fber)
5.

*Reference TAY Cohort Project 3 Protocol Paper for more information on how the data was obtained
*See Supplemental Figure 5 from the protocol paper "T1w quality Metrics (from MRIQC)" which includes signal to noise, FWHM, FBER, CJV

```{r Set up - Load csv/tsv files as data frames}

```

## Tidy data in preparation for PCA
```{r}

```


## Data exploration 
```{r}
# check if each IQM is normally distributed using histograms

```


## Principal Component Analysis
```{r}

```

